---
title: "COVID 19 Data Report analysis for UK"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  
#The YAML header sets the document title, date (generated dynamically using Sys.Date()), and output format (PDF and HTML).
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Dataset Description

This rich dataset serves as a valuable resource for understanding the global landscape of COVID-19. Compiled and maintained by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE), it offers daily time series summaries, meticulously tracking confirmed cases, deaths, and recoveries since January 21st, 2020.

The data originates from a diverse range of credible sources, including the World Health Organization (WHO), the Los Angeles Times, and QQ News. This collaborative effort ensured a comprehensive picture of the pandemic's progression across the globe.

However, it's important to note that data collection by the Johns Hopkins Coronavirus Resource Center concluded on March 10th, 2023. This cessation marks a significant shift in how we track and monitor the pandemic.

For those seeking to delve deeper into this dataset, a wealth of information is readily available on the official Johns Hopkins Github repository: link to repository <https://github.com/CSSEGISandData/COVID-19>.

## Step 0: Import Packages

```{r import_package, message = FALSE}
library(tidyverse)
library(forecast)
#tidyverse provides a suite of data manipulation tools.
#forecast is used for time series analysis and forecasting.
```

## Step 1: Import the Data

- Copy the link address of the csv file.

```{r get_data}
# Define the base URL for COVID-19 data
url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/" 


# Specify filenames for confirmed and death cases
file_names = c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv")

# Combine URL and filenames to create download links
urls <- str_c(url_in, file_names)

# Print the download links for reference
cat("Download URLs:\n", urls, "\n")

```
## Import Description
The code defines the base URL for the COVID-19 data repository.
It specifies filenames for both confirmed cases and death cases.
str_c() is used to concatenate the URL and filenames, creating download links.
read_csv() reads the confirmed cases data from the first URL.


- Use `read_csv()` to read in the data.

```{r read_global_cases, message = FALSE}

# Read confirmed cases data
global_cases <- read_csv(urls[1])
```
```{r read_global_deaths, message = FALSE}
# Read death cases data (similar to confirmed cases)
global_deaths <- read_csv(urls[2])
```
```{r head_global_cases_deaths, echo = FALSE}
# View the first few rows of confirmed and death cases data (optional)
head(global_cases)
head(global_deaths)
```

## Step 2: Tidy and Transform the Data

### **1. Tidy the columns**
  - Put each variable (**date**, **cases**, and **deaths**) in their own column.
  
  - Remove columns: **Lat** and **Long**.
  
  - Rename columns: **Province/State** and **Country/Region**.
  
  - Convert column **date** to date object.
  
```{r tidy_columns}
# Use `pivot_longer()` to make each date on a separate row
tidy_cases = global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", 
               values_to = "cases")

tidy_deaths = global_deaths %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", 
               values_to = "deaths")

```
```{r join_datasets, message = FALSE}
# Combine confirmed and death cases data using full_join()
global_tidy <- tidy_cases %>%
  full_join(tidy_deaths) %>%
  select(-c(Lat, Long)) %>%  # Remove unnecessary columns
  rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>%
  mutate(date = mdy(date))   # Convert date format

# View the first few rows of the tidied data (optional)
head(global_tidy)
```

### **2. Tidy the rows**
  - Filter the rows of **Country_Region** of Taiwan*.
  
```{r tidy_rows}
# Filter data for the United Kingdom (modify for Taiwan if needed)
uk_tidy <- global_tidy %>%
  filter(Country_Region == "United Kingdom") %>%
  select(-Province_State)

# Summarize the tidied data for the UK
summary(uk_tidy)

```



## Step 3: Add Visualizations and Analysis

### Question 1:  What are the trends for daily cumulative confirmed cases and new confirmed cases of COVID-19 in the United Kingdom?

- Since the spread of the COVID-19 pandemic in **January 2020**, United Kingdom has seen a significant surge in confirmed cases during **Jan     2022**.

- After the outbreak of a large-scale epidemic, there were higher cases of infections in **July 2021** and **July 2022**, but there has been an overall downward trend.

```{r ggplot_1}

# Calculate new cases
uk_tidy <- uk_tidy %>%
  mutate(new_cases = cases - lag(cases)) %>%
  mutate(new_cases = ifelse(is.na(new_cases) | new_cases < 0, 0, new_cases))

# Convert cases and new cases to thousands for better visualization
uk_tidy$cases_100k <- uk_tidy$cases / 100000
uk_tidy$new_cases_k <- uk_tidy$new_cases / 1000

# Create the plot
ggplot(uk_tidy, aes(x = date)) +
  geom_line(aes(y = cases_100k, color = "Cumulative Cases"), linewidth = 0.7) +
  geom_bar(aes(y = new_cases_k, fill = "New Cases"), stat = "identity", alpha = 0.5) +
  labs(x = "Date", y = "Number of Cases") +
  scale_color_manual(values = c("Cumulative Cases" = "blue")) +
  scale_fill_manual(values = c("New Cases" = "orange")) +
  ggtitle("COVID-19 Cases in the United Kingdom") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y.right = element_text(color = "blue"),
        axis.text.y.left = element_text(color = "orange"),
        axis.title.y.left = element_text(color = "black")) +
  scale_y_continuous(
    sec.axis = sec_axis(~.*100000, labels = scales::unit_format(unit = "M", scale = 1e-6)),
    limits = c(0, max(uk_tidy$cases_100k, na.rm = TRUE) * 1.2),  # Adjust limits based on data
    breaks = seq(0, max(uk_tidy$cases_100k, na.rm = TRUE) * 1.2, by = 50), labels = function(x) paste0(x, "K")
  ) +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "2 month") +
  guides(color = guide_legend(title = NULL), fill = guide_legend(title = NULL))

```

The code calculates new cases for the UK, handles potential NA or negative values, and converts cases and new cases to thousands for better visualization.
It creates a line chart for cumulative cases and a bar chart for new cases, with appropriate labels and formatting.

### Question 2: What are the trends for daily cumulative deaths and new confirmed cases of COVID-19 in the United Kingdom?

- The number of deaths has been increasing alongside the rising number of confirmed cases.

```{r ggplot_2}
# Convert cumulative deaths to thousands
uk_tidy$deaths_k <- uk_tidy$deaths / 1000

# Create the plot
ggplot(uk_tidy, aes(x = date)) +
  geom_line(aes(y = deaths_k, color = "Cumulative Deaths"), linewidth = 0.7) +
  geom_bar(aes(y = new_cases_k, fill = "New Cases"), stat = "identity", alpha = 0.5) +
  labs(x = "Date", y = " ") +
  scale_color_manual(values = c("Cumulative Deaths" = "purple")) +
  scale_fill_manual(values = c("New Cases" = "orange")) +
  ggtitle("COVID-19 Cases and Deaths in the United Kingdom") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y.right = element_text(color = "purple"),
        axis.text.y.left = element_text(color = "orange")) +
  scale_y_continuous(
    sec.axis = sec_axis(~., labels = function(x) paste0(x, "K")),
    limits = c(0, max(uk_tidy$deaths_k, na.rm = TRUE) * 1.2),  # Adjust limits based on data
    breaks = seq(0, max(uk_tidy$deaths_k, na.rm = TRUE) * 1.2, by = 50), labels = function(x) paste0(x, "K")
  ) +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "2 month") +
  guides(color = guide_legend(title = NULL), fill = guide_legend(title = NULL))

```

The code converts cumulative deaths to thousands for better visualization.
It creates a line chart for cumulative deaths and a bar chart for new cases, with appropriate labels and formatting.

### Question 3: Can we predict the future number of confirmed cases in the United Kingdom?

- **Purpose**: Predict the future number of COVID-19 confirmed cases in United Kingdom for the upcoming year based on the data collected by JHU CSSE.

- **Methods**: Use a **ARIMA model** to model and forecast.
  - Use `auto.arima()` to build a time series model.
  - Use `forecast()` to predict future data.

```{r model}
# Use `ts()` convert data into a time series object
ts_cases = ts(uk_tidy$cases)

# ARIMA model
arima_model = auto.arima(ts_cases)

# Make predictions using the established ARIMA model
future_forecast = forecast(arima_model, h = 365)

# Create a chart
plot(future_forecast, main = "United Kingdom COVID-19 Cases Forecast", yaxt = "n", xaxt = "n")
grid(lty = "dotted", col = "gray")

# Draw the y-axis labels
y_labels = c(-10, 0, 10, 20, 30, 40) * 1e6
axis(2, at = y_labels, labels = paste0(y_labels / 1e6, "M"))

# Draw the x-axis labels
x_labels = c(0, 500, 1000, 1500)
x_labels_dates = c(uk_tidy$date[1], uk_tidy$date[1] + 499, uk_tidy$date[1] + 999, 
                   uk_tidy$date[1] + 1499)
axis(1, at = x_labels, labels = paste0(x_labels_dates))

# Add text to x-axis and y-axis
mtext("Date", side = 1, line = 3)
mtext("Number of Cases", side = 2, line = 3)


```

  -  *Data Filtering:* We filter the global dataset to include only data for the United Kingdom.
  -  *Data Preparation:* Calculate new cases and convert cases and new cases to thousands for better visualization.
  -  *Time Series Creation:* Create a time series object for total cases in the UK.
  -  *Model Fitting:* Fit an ARIMA model to the time series data.
  -  *Forecasting:* Generate forecasts for the next 30 days.
  -  *Visualization:* Plot the forecast for visual inspection.

  - The code creates a time series object (uk_ts_cases) for the total cases in the UK, specifying the start date and frequency.
  - An ARIMA model is fitted to the time series data using auto.arima().
  - The forecast() function generates predictions for the next 30 days based on the fitted model.
  - The plot() function visualizes the forecast.

**Additional Analysis**

  - *Error Analysis:*   Calculate the accuracy of the forecast using metrics like Mean Absolute Error (MAE), Mean Squared Error (MSE), and Root Mean     Squared Error (RMSE).Compare the forecast with actual data (if available) to assess its performance.

  - *Model Improvement:*  Explore different time series models (e.g., SARIMA, exponential smoothing) to find the best fit for the data.
    Consider incorporating external factors (e.g., government policies, public health interventions) into the model for improved accuracy.

  - *Visualization:* Create a plot comparing the actual and forecasted values to visually assess the model's performance.
    Note: To calculate forecast accuracy metrics, we'll need actual data for the forecast period.So skipping it for now. 

## Step 4: Add Bias Identification

Bias Identification for the United Kingdom COVID-19 Analysis

**1. Personal Bias**
  
  - *Before analysis*: Given the extensive media coverage of the UK's initial COVID-19 wave, there might be a tendency to overemphasize the early        impact while overlooking subsequent trends.
  - *After analysis*: Overreliance on specific data points or visualizations could lead to a narrow interpretation, neglecting potential alternative     explanations for observed patterns.

**2. Other Bias**
  
  - *Reporting Bias*: The UK's complex healthcare system, with devolved responsibilities, might introduce variations in data collection and reporting     practices across different regions. This could affect data consistency and comparability.
  - *Temporal Bias*: The UK's experience with COVID-19 has been marked by distinct phases, including lockdowns, reopenings, and new variants.            Analyzing data without considering these temporal factors might lead to misleading conclusions.
  - *Policy Bias*: The impact of government policies, such as lockdown measures, vaccination campaigns, and testing strategies, can significantly        influence the data. Overlooking these factors could bias the analysis.


## Conclusion

This analysis provides a comprehensive overview of COVID-19 trends in the United Kingdom, including visualizations, time series modeling, and forecasting. The results can be used to understand the past, assess the present, and make predictions about the future trajectory of the pandemic. Further analysis and refinement of the models can provide more accurate and informative insights.